---
title: "Making a Prediction Map using leaflet()"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
  pdf_document: default
---

This document outlines how to plot Maxent predictions onto a `leaflet` map --- which differs from the `ggplot` version in that it displays the predictions over a base map. Most of this was covered by Jeremy in class on 3/12, so all the code until the leaflet section should be what was covered in class.

## Install and Load Packages

These packages will be necessary for performing the code that follows, so here we will install (if not done already) and load them:

### Installation (only needs to be done once)

```{r, eval=FALSE}
install.packages("dismo")
install.packages("dplyr")
install.packages("geodata")
install.packages("rJava")
install.packages("ggplot2")
install.packages("leaflet")
install.packages("maps")
install.packages("mapview")
install.packages("webshot2")
```

### Load libraries (needs to be done every session)

```{r, results='hide', warning=FALSE, message=FALSE}
library(dismo)
library(dplyr)
library(geodata)
library(rJava)
library(ggplot2)
library(leaflet)
library(maps)
library(mapview)
library(webshot2)
```

## Predictions from Maxent

### Current SDM

I believe the only difference between the current and future SDMs is which climate (WorldClim or CMIP6) is being used, so I will only put the current SDM instructions in this document.

1.  Get occurrence data

```{r, message=FALSE, warning=FALSE}
# Get latitude and longitude from our species data
occurrenceCoords<-read.csv("data/cleanedData.csv") %>%
  dplyr::select(decimalLongitude, decimalLatitude)

# Convert to spatial points, necessary for modelling and mapping
occurrenceSpatialPts <- SpatialPoints(occurrenceCoords, 
                                      proj4string = CRS("+proj=longlat"))
# Now get the climate data

# Make sure RAM is bumped up (I'm using 16GB)

# This command downloads 19 different raster files, one for each climate variable
# raster: image file comprised of pixels (locations), with each pixel associated with some value (e.g. 25 degrees)

# worldclim_global(var = "bio", res = 2.5, path = "data/", version = "2.1")

# Update .gitignore to prevent huge files getting pushed to github!
# Add: data/wc2.1_2.5m/

```

Here are the meanings of the bioclimatic variables (bio1 to bio19) provided by WorldClim:

-   bio1: Mean annual temperature

-   bio2: Mean diurnal range (mean of monthly (max temp - min temp))

-   bio3: Isothermality (bio2/bio7) (\*100)

-   bio4: Temperature seasonality (standard deviation \*100)

-   bio5: Max temperature of warmest month

-   bio6: Min temperature of coldest month

-   bio7: Temperature annual range (bio5-bio6)

-   bio8: Mean temperature of wettest quarter

-   bio9: Mean temperature of driest quarter

-   bio10: Mean temperature of warmest quarter

-   bio11: Mean temperature of coldest quarter

-   bio12: Annual precipitation

-   bio13: Precipitation of wettest month

-   bio14: Precipitation of driest month

-   bio15: Precipitation seasonality (coefficient of variation)

-   bio16: Precipitation of wettest quarter

-   bio17: Precipitation of driest quarter

-   bio18: Precipitation of warmest quarter

-   bio19: Precipitation of coldest quarter

```{r, message=FALSE, warning=FALSE}
# We need to feed the model a "raster stack", or a bunch of rasters laid on top of each other.

# Let's make a list of the raster files:
climList <- list.files(path = "data/wc2.1_2.5m/", 
                       pattern = ".tif$", 
                       full.names = T)

# Stacking the bioclim variables to process them at one go
# Each "layer" of the stack is a file for one environmental variable
currentClimRasterStack <- raster::stack(climList)

# Show one environmental layer ( = annual precepitation)
plot(currentClimRasterStack[[1]]) 
plot(occurrenceSpatialPts, add = TRUE) 

```

Looks good, we can see where our data is.

2.  Create pseudo-absence points

```{r, message=FALSE, warning=FALSE}
# First we need a raster layer to make the points up on, just picking 1
# mask: raster object that determines the area where we are generating pts
mask <- raster(climList[[1]]) 

# Determine geographic extent of our data 
# (so we generate random points reasonably nearby)
geographicExtent <- extent(x = occurrenceSpatialPts)

# Random points for background (same number as our observed points we will use)
# Seed set so we get the same background points each time we run this code
set.seed(45)

# When you set the seed to a specific value, 
# R will generate the same sequence of random numbers every time you run your code, 
# provided that you use the same seed value. This is useful for debugging, 
# sharing code, and ensuring that others can reproduce your results exactly.

backgroundPoints <- randomPoints(mask = mask, 
                                 n = nrow(occurrenceCoords), # same n 
                                 ext = geographicExtent, 
                                 extf = 1.25, # draw a slightly larger area 
                                 warn = 0) 

# Add col names (can click and see right now they are x and y)
colnames(backgroundPoints) <- c("longitude", "latitude")
```

3.  Convert occurrence and environmental data into format for model

```{r, message=FALSE, warning=FALSE}
# Data for observation sites (presence and background), with climate data
# Creates grids of climate measurements, per point
occEnv <- na.omit(raster::extract(x = currentClimRasterStack, y = occurrenceCoords))

# Same for absence
absenceEnv<- na.omit(raster::extract(x = currentClimRasterStack, y = backgroundPoints))

# (0 = abs, 1 = pres)
# Bunch of true/false... we'll use later
presenceAbsenceV <- c(rep(1, nrow(occEnv)), rep(0, nrow(absenceEnv))) 

# Create a single data frame with both presence and absence points for model training
presenceAbsenceEnvDf <- as.data.frame(rbind(occEnv, absenceEnv))
```

4.  Current SDM with dismo::maxent

```{r}
# Create a new folder called maxent_outputs
# If you get a Java error, restart R, and reload the packages
habronattusCurrentSDM <- 
  dismo::maxent(x = presenceAbsenceEnvDf,    # env conditions
                p = presenceAbsenceV,         # 1:presence or 0:absence
                path = paste("maxent_outputs"))     # maxent output dir 
```

## Plotting with `ggplot`

The ggplot version requires the results to be turned into a dataframe object first.

```{r}
# currentClimRasterStack is huge and it isn't reasonable to predict over 
# whole world
# First we will make it smaller

# Choose here what is reasonable for your pts (where you got background pts from)
predictExtent <- 1.25 * geographicExtent
geographicArea <- crop(currentClimRasterStack, predictExtent, snap = "in")

# Crop currentClimRasterStack to the extent of the map you want
# Makes a RasterLayer with a prediction based on a a fitted model object.
habronattusPredictPlot <- raster::predict(habronattusCurrentSDM, geographicArea) 

# For ggplot, we need the prediction to be a data frame 
raster.spdf <- as(habronattusPredictPlot, "SpatialPixelsDataFrame")
habronattusPredictDf <- as.data.frame(raster.spdf)
```

Now we can plot it!

```{r}
# Plot in ggplot
wrld <- ggplot2::map_data("world")

# Determine the bounding box of the map
xmax <- max(habronattusPredictDf$x)
xmin <- min(habronattusPredictDf$x)
ymax <- max(habronattusPredictDf$y)
ymin <- min(habronattusPredictDf$y)

# Using ggplot, b/c I couldn't get it to work in leaflet :(
ggplot() +
  geom_polygon(data = wrld, mapping = aes(x = long, y = lat, group = group),
               fill = "grey75") +
  geom_raster(data = habronattusPredictDf, aes(x = x, y = y, fill = layer)) + 
  scale_fill_gradientn(colors = terrain.colors(10, rev = T)) +
  coord_fixed(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = F) +
  scale_size_area() +
  borders("state") +
  borders("world", colour = "black", fill = NA) + 
  labs(title = bquote(paste(bold("SDM of"), 
                            bolditalic(" Habronattus americanus"),
                            bold(" Under Current Climate Conditions"))),
       x = "Longitude",
       y = "Latitude",
       fill = "Environmental Suitability")+ 
  theme(legend.box.background = element_rect(),
        legend.box.margin = margin(5,5,5,5)) 
  
# Save the file
# Dimensions may require some tinkering
ggsave("output/habronattusCurrentSdm.jpg", width = 35, height = 25, units = "cm")
```

## Plotting with `leaflet`

Using leaflet requires the raster result prior to conversion in addition to the data frame version. Here I will present two options for mapping.

1.  No modifications to suitability results:

```{r}
# Color palette for legend
pal <- colorBin(palette = c("#4A12A4", "#FFD900"), 
                    domain = habronattusPredictDf$layer,
                    bins = 10)

# Plot map
map1 <-
leaflet(options = leafletOptions(zoomControl = FALSE, # Removes zoom button
                                 attributionControl = FALSE, # Removes banner
                                 maxZoom = 19)) %>% 
  # Add basemap
  addProviderTiles("OpenStreetMap.Mapnik") %>%
  # Add Maxent results
  addRasterImage(habronattusPredictPlot, 
                 colors = viridis::plasma(99),
                 opacity = 0.80) %>%
  # Add legend
  addLegend(position = "bottomleft",
            pal = pal,
            values = habronattusPredictDf$layer,
            title = "Suitability",
            opacity = 1)

map1

# Save this map
# mapshot2(map1, file = "output/habro_current_sdm_version1.png")
```

2.  Limiting result presentation to those only above 0.1 suitability:

    For better visiblity of the map under the results, we can choose to omit suitability values that are below 0.10. ***Pay attention to whether the raster or data frame version of our results is being used***.

```{r}
# We'll use a copy of the original in case we ever decide we don't want to do that
habronattusPredictPlot_mod <- habronattusPredictPlot

# Then we remove the values below 0.1
habronattusPredictPlot_mod[habronattusPredictPlot_mod < 0.10] <- NA

# Convert this to a data frame (for the legend)
raster.spdf2 <- as(habronattusPredictPlot_mod, "SpatialPixelsDataFrame")
habronattusPredictDf_mod <- as.data.frame(raster.spdf2)

# Color palette for legend
pal2 <- colorBin(palette = c("#4A12A4", "#FFD900"), 
                 domain = habronattusPredictDf_mod$layer,
                 bins = 10)

# Plot map
map2 <-
leaflet(options = leafletOptions(zoomControl = FALSE, # Removes zoom button
                                 attributionControl = FALSE, # Removes banner
                                 maxZoom = 19)) %>% 
  # Add basemap
  addProviderTiles("OpenStreetMap.Mapnik") %>%
  # Add Maxent results
  addRasterImage(habronattusPredictPlot_mod, 
                 colors = viridis::plasma(99),
                 opacity = 0.80) %>%
  # Add legend
  addLegend(position = "bottomleft",
            pal = pal2,
            values = habronattusPredictDf_mod$layer,
            title = "Suitability",
            opacity = 1)

map2

# Save this map
# mapshot2(map1, file = "output/habro_current_sdm_version2.png")
```
